---
title: "Examine and Filter Cells and Genes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Read in data matrix
```{r} 
data = read.table("Glioblastoma_expressed_genes.txt", header=T, row.names=1)  ## CHANGE TO YOUR INPUT MATRIX

```

# Examine distributions of counts of genes and cells
```{r}
reads_per_cell = colSums(data)
reads_per_gene = rowSums(data)
genes_per_cell = colSums(data>0)
cells_per_gene = rowSums(data>0)

hist(log10(reads_per_cell+1),main='reads per cell',col='wheat')
hist(log10(genes_per_cell+1), main='genes per cell', col='wheat')
plot(reads_per_cell, genes_per_cell, log='xy', col='wheat')
hist(log10(reads_per_gene+1),main='reads per gene',col='wheat')
```

```{r}
plot(sort(genes_per_cell), xlab='cell', log='y', main='genes per cell (ordered)')
```
#Cell filtering criteria: define min and max genes per cell

```{r}
##################################################
## ********* USER DEFINED SECTION ***************
##################################################

#  set upper and lower thresholds for genes per cell:
MIN_GENES_PER_CELL = 350  ## user-defined setting
MAX_GENES_PER_CELL = 1800  ## user-defined setting

# now replot with the thresholds being shown:
plot(sort(genes_per_cell), xlab='cell', log='y', main='genes per cell (ordered)')
abline(h=MIN_GENES_PER_CELL, col='green')  # lower threshold
abline(h=MAX_GENES_PER_CELL, col='green') # upper threshold
```


# Examine percent mitochondrial read content
```{r}
# define the mitochondrial genes
mito_genes = grep("^mt-", rownames(data) , ignore.case=T, value=T)
print(mito_genes)
```

```{r}
# compute pct mito
mito_gene_read_counts = colSums(data[mito_genes,])
pct_mito = mito_gene_read_counts / reads_per_cell * 100
plot(sort(pct_mito))
```

# Decide on maximum allowed percent mitochondrial reads:
```{r}
##################################################
## ********* USER DEFINED SECTION ***************
##################################################

MAX_PCT_MITO = 10   ## user-defined setting

plot(sort(pct_mito))
abline(h=MAX_PCT_MITO, col='red')
```


# cell selection as per Peter Karchenko - the Pagoda way

```{r}
df = data.frame(reads_per_cell=reads_per_cell, genes_per_cell=genes_per_cell)
head(df)
```


# Plot gene_per_cell vs. reads_per_cell, define outliers

```{r}
library(MASS)
df = df[order(df$reads_per_cell),] # order by reads_per_cell
plot(df, log='xy')
m <- rlm(genes_per_cell~reads_per_cell,data=df) # robust linear model, not sens to outliers
p.level = 1e-3
# predict genes_per_cell based on observed reads_per_cell
suppressWarnings(pb <- data.frame(predict(m, interval='prediction', 
                                          level = 1-p.level, # define conf interval
                                          type="response")))
polygon(c(df$reads_per_cell, rev(df$reads_per_cell)),
        c(pb$lwr, rev(pb$upr)), col=adjustcolor(2,alpha=0.1), border = NA)

# identifier outliers as having observed genes_per_cell outside the prediction confidence interval
outliers <- rownames(df)[df$genes_per_cell > pb$upr | df$genes_per_cell < pb$lwr];
points(df[outliers,],col=2,cex=0.6)
```

# Before pruning cells, let's make a backup copy of the original matrix:
```{r}
data.prefiltered = data
```

# Now, let's do some pruning to remove 'bad' cells
```{r}
filtered_data = data.prefiltered # just in case we re-run this block using different thresholds.

###############################################################
# prune genes, require a gene to be expressed in at least 3 cells

filtered_data.prefiltered = filtered_data
filtered_data = filtered_data[cells_per_gene >= 3,]  ## user can change this if needed.

###############################################################
# prune cells
valid_cells = colnames(filtered_data) # all cells
message('starting with: ', length(valid_cells), ' cells') # number starting with

## remove cells based on gene count criteria:
valid_cells = valid_cells[genes_per_cell >= MIN_GENES_PER_CELL & genes_per_cell <= MAX_GENES_PER_CELL]  # set values based on your evaluation above
message('after filtering low and high gene count outliers: ', length(valid_cells), ' cells') # number after filtering based gene count thresholds

## remove cells having excessive mito read content
valid_cells = valid_cells[valid_cells %in% names(pct_mito)[pct_mito <= MAX_PCT_MITO]]
message('after removing high-mito cells: ', length(valid_cells), ' cells') # number remaining after high-mito cells removed

## remove cells identified as outliers via the Karchenko method
valid_cells = valid_cells[ ! valid_cells %in% outliers]
message('after removing final outliers: ', length(valid_cells), ' cells') # number surviving outlier detection

## update the count matrix to contain only the valid cells
filtered_data = filtered_data[,valid_cells]

write.table(filtered_data, file="filtered_data.counts.matrix", quote=F, sep="\t")
```



